---
title: "Final Project EDA"
author: "XI KANG, MINGZE YAN"
date: "4/25/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  message = FALSE, warning = FALSE)
```

```{r load-packages-set-seed}
## Load Packages, Set Seed
library(tidyverse)
library(tidymodels)
library(skimr)
library(naniar)
library(kableExtra)
library(corrplot)
set.seed(123)
```

## Introduction

This is the EDA section for the Stat 301-3 final project. The goal of this project is to predict one's risk for stroke through a classification approach. 
<br>

Based on the information from the [World Health Organization (WHO)](https://dx.doi.org/10.2471/BLT.16.181636), stroke is the second leading cause of death and the third leading cause for disability. Nevertheless, effective prevention actions can be applied to reduce the fatal effects of this medical emergency and lower the risk for mortality and disability. Thus, to provide targeted support and effective prevention strategies, it is crucial to identify the risks for stroke based on common factors related to one's lifestyle or chronic health conditions. 
<br>

Thus, driven by the incentives mentioned above, our project focuses on predicting the risk for stroke based on input parameters about one's lifestyle and health status. The research question of this project is a predictive classification question. 
<br>

## Initial Overview

### Data Source

The dataset used by this project is the "Stroke Prediction Dataset" provided on [Kaggle](https://www.kaggle.com/fedesoriano/stroke-prediction-dataset). We first loaded in the raw dataset and cleaned the names.  The initial data tidying process it completed in the FinalProject_tidy_data.r file. The code used is presented below:
```{r tidy-data, eval=FALSE}
## Load raw dataset ----
stroke_dat <- 
  read_csv("data/unprocessed/healthcare-dataset-stroke-data.csv") %>% 
  clean_names() %>% 
  mutate(
    gender = factor(gender), 
    ever_married = factor(ever_married),
    work_type = factor(work_type),
    residence_type = factor(residence_type),
    smoking_status = factor(smoking_status),
    stroke = factor(stroke, levels = c(1, 0), 
                    labels = c("Yes", "No")),
    hypertension = factor(hypertension, levels = c(1, 0), 
                          labels = c("Yes", "No")), 
    heart_disease = factor(heart_disease, levels = c(1, 0), 
                           labels = c("Yes", "No")), 
    bmi = as.numeric(bmi)
  ) %>% 
  select(-id)

## initial skimming ----
skim_without_charts(stroke_dat)

## save processed data ----
stroke_dat %>% 
  write_rds("data/processed/stroke_data.rds")
```
<br>

### Dataset Overview

Using methods from `skimr`, we noticed that the original dataset contains 5110 rows and 12 columns. For the convenience of the later processes, we first turned the categorical variables `gender`, `ever_married`, `work_type`, `residence_type`, and `smoking_status` into factors. The outcome variable `stroke` and variables `hypertension` and `heart_disease` were initially identified as numeric variables, but they should be categorical. Thus, we turned them into a factors. Since `id` will not be used as a categorical variable, we removed it from the original dataset. 
<br>

After initial tidying, we loaded in the processed data and performed a train-test split, stratified by the outcome variable `stroke`. We used 70% of the data for training and 30% for testing: 
```{r, load-data-and-split}
stroke_data <- readRDS("data/processed/stroke_data.rds")

## split data ----
stroke_split <- initial_split(stroke_data, prop = 0.7, strata = stroke)
stroke_train <- training(stroke_split)
stroke_test <- testing(stroke_split)
```
<br>

The EDA section will be performed using the training set data, `stroke_train`.
<br>

After the split, we first obtained an overview on the training set using the methods from the `skimr` package. We noticed that the training set contains 11 variables, and 3577 observations. Among them 8 variables are factors and 3 variables are numeric. The outcome variable `stroke` is a categorical variable of type factor. 
<br>

### Missing Data

During the initial skimming, we noticed that only one variable `bmi` contains missing values. To obtain a wholistic overview on the general situation of missingness in the training data, we created a plot showing the amount of missing data in each column using the `vis_miss()` function from `naniar`: 
```{r, fig.height = 7, fig.width = 7}
# overview of missingness
stroke_train %>% 
  vis_miss(sort_miss = TRUE)
```
<br>

As shown, within the 11 columns in the dataset, there is only column `bmi` with missing data. The response variable `stroke` does not have any missing values. To clearly present the percentage of missingness for each variable, we created the table below using `miss_var_summary()`, and we highlighted the rows for the variable with missing value in blue:  
```{r}
# overview table
stroke_train %>% 
  miss_var_summary() %>% 
  kbl() %>% 
  kable_classic() %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  row_spec(1, color = "white",
           background = "blue") %>% 
  kableExtra::footnote(
    general = "Number and Percentage of Missing Values in Each Variable"
  )
```
<br>

As shown, `bmi` is the only variable with missing values and its percentage of missingness is about 4%. Since the degree of missingness is very low, we plan to impute `bmi` using bagged tree imputation. 
<br>

## Essential Findings

### Outcome variable

We first performed a univariate investigation of the response variable `stroke`. The barchart below provides an overview on the distribution of `stroke` in the training set: 
```{r outcome-var-plot, fig.height= 7, fig.width=7}

# plot
stroke_train %>%
  ggplot(aes(stroke, fill = stroke)) + 
  geom_bar(show.legend = FALSE) + 
  labs(
    x = "Stroke", 
    y = "Number of Observations", 
    title = "Distribution of Response Variable"
  ) + 
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25
  )
```
<br>

As shown above, even though we stratified by the outcome variable `stroke` when performing data splitting, there is still severe class imbalance in the outcome variable. There are 180 observations in level `Yes` and 3397 observations in level `No`. Depending on the plans for resampling, we might need to consider using over-sampling or under-sampling to avoid errors in the resampling and cross-validation process. Further decisions will be made after consulting the professors and TAs. 
<br>

### Numeric Predictors

#### Correlation among Predictors

We used a correlation plot to explore the correlation among numeric predictors. Since the degree of missingness is very low for `bmi` we temporarily dropped the `NA` in plotting to avoid errors.  
```{r, fig.height= 8, fig.width=8}
# correlation plot
stroke_train %>% 
  select(age, avg_glucose_level, bmi) %>% 
  
  drop_na() %>% 
  # compute correlation matrix
  cor() %>% 
  # visualize
  corrplot(type = "upper", 
           title = "Correlations between Numeric Variables", 
           mar = c(0, 0, 1, 0))
```
<br>

As shown, there is no near-perfect collinearity between any pair of number predictors. All three variables are positively related to each other. A table showing the correlation between the numeric predictors is also presented below: 
```{r}
# correlation between predictors
stroke_train %>% 
  select(age, avg_glucose_level, bmi) %>% 
  
  drop_na() %>% 
  # compute correlation matrix
  correlate() %>% 
  # turn into a tibble
  stretch() %>% 
  rename("correlation" = "r") %>% 
  arrange(desc(correlation)) %>%
  # temporary var `row_id` to help removing repeated info
  mutate(row_id = row_number()) %>% 
  # filter out even rows
  drop_na() %>% 
  filter(row_id %% 2 == 0) %>% 
  # remove temporary var
  select(-row_id) %>%
  kbl() %>% 
  kable_classic()
```
<br>

As shown, the highest correlation coefficient is between variables `bmi` and `age`, which is 0.3461315. The table comfirms that there is no strong linear correlation between the numeric predictors.
<br>

#### `age` predictor

In common knowledge, age is often closely related to one's risk for stroke. Also, knowing the distribution of variable `age` in this dataset can help us understand the age range of population for which the result of this project is applicable. Thus, we plotted the distribution of `age` in the training set: 
```{r}
# distribution of age
ggplot(stroke_train, aes(age)) + 
  geom_histogram(binwidth = 1) + 
  labs(
    x = "age", 
    y = "number of observations", 
    title = "Distribution of `age` in the Training Set"
  )
```
<br>

As shown, the distribution of `age` is about normal. Also, the observations range from 0 to 80-year old without any gaps in between. There are two abrupt peaks: one slightly below 10 and one slightly above 75, suggesting that there are more observations with these two age values. However, no obvious outliers are observed. 
<br>

We then explored the relation between `age` and the outcome variable `stroke` using the boxplot below: 
```{r}
# relation between `age` and `stroke`
ggplot(stroke_train, aes(age, stroke)) + 
  geom_boxplot() + 
  labs(
     x = "Age",
     y = "Stroke",
     title = "The relationship between whether to have stroke and age"
  )
```
<br>

As shown, there is a clear correlation between age and an individual's chance of getting stroke. In the training set, almost all observations with value `Yes` for `stroke` have age above 35, with the exception of one outlier at around 17. Also, the median for the `age` value for observations with `No` for `stroke` is just above 40, much lower than the median for the observations with `Yes` for `stroke` (above 70). It can be implied that, in the dataset used, individuals who had stroke tend to be older in age. 

#### Skewness in `avg_glucose_level` and `bmi`

We also explored the distribution of `avg_glucose_level` and `bmi` in the training set: 
```{r}
# distribution of avg_glucose_level
ggplot(stroke_train, aes(avg_glucose_level)) + 
  geom_histogram(binwidth = 5) + 
  labs(
    x = "Average Glucose Level", 
    y = "number of observations", 
    title = "Distribution of `avg_glucose_level` in the Training Set"
  )
```
<br>

As shown, the distribution of `avg_glucose_level` is severely right-skewed. With the number of observations decreasing abruptly above 110. This result suggests that there are fewer observations with `avg_glucose_level` above 110. This result makes sense, considering the online resources suggest that a normal blood glucose level ranges from 80 to 140. An blood glucose level in the range 140-200 should be considered as pre-diabetes. Thus, we can imply from the plot that,in the training set, there are more observations with normal blood sugar level than observations with pre-diabetes conditions. 
<br>


```{r}
# distribution of bmi
ggplot(stroke_train, aes(bmi)) + 
  geom_histogram(binwidth = 1) + 
  labs(
    x = "BMI (Body Mass Index)", 
    y = "number of observations", 
    title = "Distribution of `bmi` in the Training Set"
  )
```
<br>

As shown, the distribution of `bmi` in the training set is also severly right-skewed. There are near-zero observations for `bmi` values above 50 and values below 10. From online resources, we learned that the normal range for BMI value is 18.5 to 25. BMI values less than 18.5 suggest underweight. Values from 25.0 to 30 suggests overweight range, and BMI above 30 suggests obesity. Thus, it makes sense for the distribution of `bmi` values to peak around the range from 25 to 30. 
<br>

Based on the observed skewness, we decided to log-transform variables `avg_glucose_level` and `bmi` in creating the recipes later. 
<br>



### Categorical Predictors
#### Distribution of `gender`
```{r}
ggplot(stroke_train, aes(x = gender)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25,
    hjust = 0.5
  ) +
  labs(
    title = "Distribution of gender"
  ) +
  theme_minimal()
```
The distribution of gender shows that there are around 600 more female patients in the training data than male ones and only one patient classifying itself as `Other` gender.

#### Distribution of `ever_married` and `smoking_status`
```{r}
ggplot(stroke_train, aes(x = ever_married)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25,
    hjust = 0.5
  ) +
  labs(
    title = "Distribution of ever_married"
  ) +
  theme_minimal()
```
The graph shows that there are approximately 1100 more patients having married before. 

```{r}
ggplot(stroke_train, aes(x = smoking_status)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25,
    hjust = 0.5
  ) +
  labs(
    title = "Distribution of smoking_status"
  ) +
  theme_minimal()
```
The distribution of `smoking_status` shows that patients having never smoked are the largest (1333) and patients who have unknown smoking status are the second largest (1081). Patients who smoke and formerly smoked are only 55 apart and there is no huge difference between these two types of patients.

#### Distribution of `work_type` and `residence_type`
```{r}
ggplot(stroke_train, aes(x = work_type)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25,
    hjust = 0.5
  ) +
  labs(
    title = "Distribution of work_type"
  ) +
  theme_minimal()
```
The graph shows that the number of patients having `children`, `Govt_job` types of work are pretty much the same and the number of patients who are self-employed are a little more. The number of patients who have never worked is only 17 and is the least. The number of patients who have private jobs is the largest (2047). Therefore, if patients who are self-employed have fewer proportion in the following graph, it makes sense.

```{r}
ggplot(stroke_train, aes(x = residence_type)) +
  geom_bar() +
  geom_text(
    aes(label = ..count..), 
    stat = "count",
    position = position_dodge(width = 0.9), 
    vjust = -0.25,
    hjust = 0.5
  ) +
  labs(
    title = "Distribution of residence_type"
  ) +
  theme_minimal()
```
The graph shows that the number of patients staying in urban area and rural area is pretty close with more around 80 patients more living in urban area.

#### `gender` and `stroke`
```{r}
ggplot(stroke_train, aes(x = stroke, fill = gender)) + 
  geom_bar(position = "fill") + 
  labs(
     x = "whether to have stroke",
     y = "frequency",
     title = "The relationship between whether to have stroke and gender"
  )
```
<br>

The graph shows that the proportion of male having stroke is larger than the proportion of male not having stroke and the proportion of female having stroke is smaller than proportion of female not having stroke. The proportion of male having stroke is larger than female having stroke.

#### `ever_married` and `stroke`
```{r}
ggplot(stroke_train, aes(x = stroke, fill = ever_married)) +
   geom_bar(position = "fill") +
   labs(
     x = "whether to have stroke",
     y = "frequency",
     title = "The relationship between whether to have stroke and whether ever married"
   )
```
<br>

The graph shows that the proportion of patients ever married having stroke is much higher than the proportion of patients not ever married. However, the proportion of patients ever married having stroke is  higher than the proportion of patients not ever married. It might be caused by many patients have married before. Nevertheless, the huge difference between patients ever married and not ever married having stroke indicates a potential correlation between `ever_married` and `stroke`.

#### `work_type` and `stroke`
```{r}
stroke_train_wt <- stroke_train %>%
   group_by(stroke) %>% 
   count(work_type) %>% 
   mutate(wt_prop = n / sum(n))
 
 ggplot(stroke_train_wt, aes(x = stroke, y = wt_prop, fill = work_type)) +
   geom_col(position = "dodge") +
   labs(
     x = "whether to have stroke",
     y = "proportion",
     title = "The relationship between whether to have stroke and patient work type"
   )
```
<br>

By computing each work type proportion, the graph shows that private work type has the largest proportion in both having and not having stroke. children work type has the smallest proportion in having stroke and the third largest proportion in not having stroke. There is no people having stroke who have never worked before and there is only a tiny portion of people not having stroke who have never worked before. 

#### `smoking_status` and `stroke`
```{r}
stroke_st <- stroke_train %>%
   group_by(stroke) %>% 
   count(smoking_status) %>% 
   mutate(st_prop = n / sum(n))
 
 ggplot(stroke_st, aes(x = stroke, y = st_prop, fill = smoking_status)) +
   geom_col(position = "dodge") +
   labs(
     x = "whether to have stroke",
     y = "proportion",
     title = "The relationship between whether to have stroke and patient smoking status"
   )
```
<br>

By calculating the proportion of people's smoking status in each group, the graph indicates that the proportion of people who never smoked is the largest in both people having stroke and not having stroke. Moreover, the proportion of people who have never smoked is larger for people not having stroke than having stroke. The proportion of people who have formerly smoked in having stroke is much higher than in not having stroke. The proportion of people who smokes in having stroke is a little more than in not having stroke. Finally, the proportion of people whose smoking status is unknown has the second largest proportion in not having stroke which will add some difficulty in our understanding. 

#### `work_type` and `ever_married`
```{r}
stroke_mw <- stroke_train %>%
   group_by(work_type) %>% 
   count(ever_married) %>% 
   mutate(m_prop = n / sum(n))
 
 ggplot(stroke_mw, aes(x = work_type, y = m_prop, fill = ever_married)) +
   geom_col(position = "dodge") +
   labs(
     x = "work_type",
     y = "proportion",
     title = "The relationship between patient work type\nand whether patient has ever married"
   )
```
<br>

We also want to look into the potential relationship between patient work type and whether patient has never married. Then, we observe that patients in children work type and who have never worked have never married before. All the other work types including government-related jobs, private jobs, self-employed have higher proportion in people who have married than who have never married. Therefore, it gives us somewhat strong correlation between `work_type` and `ever_married`.




## Secondary Findings
#### `residence_type` and `stroke`
```{r}
ggplot(stroke_train, aes(x = stroke, fill = residence_type)) +
   geom_bar(position = "fill") +
   labs(
     x = "whether to have stroke",
     y = "frequency",
     title = "The relationship between whether to have stroke and patient residence type"
   )
```
The graph shows that there is basically no difference between whether patient lives and whether they have stroke. Therefore, we can potentially eliminate this variable out of our predictive model.

#### `smoking_status` and `gender`
```{r}
stroke_stg <- stroke_train %>%
   group_by(gender) %>% 
   count(smoking_status) %>% 
   mutate(st_prop = n / sum(n))
 
 ggplot(stroke_stg, aes(x = gender, y = st_prop, fill = smoking_status)) +
   geom_col(position = "dodge") +
   labs(
     x = "gender",
     y = "proportion",
     title = "The relationship between patient gender and smoking status"
   )
```
Then, we find out proportions of people's smoking status in each gender and the graph doesn't show an obvious correlation between gender and each group of smoking_status. It can be noticed that the proportion of female who never smoked is higher than the proportion of male who never smoked. It is also worth noting that Other gender has formerly smoked because there is only one person whose gender is Other. 

#### `work_type` and `residence_type`
```{r}
 stroke_wtr <- stroke_train %>%
   group_by(work_type) %>% 
   count(residence_type) %>% 
   mutate(rt_prop = n / sum(n))
 
 ggplot(stroke_wtr, aes(x = work_type, y = rt_prop, fill = residence_type)) +
   geom_col(position = "dodge") +
   labs(
     x = "work_type",
     y = "proportion",
     title = "The relationship between patient work type and residence type"
   )
```
We also compute proportions of residence type in each work type and it can be seen that there is no strong correlation between whether patient lives and its work type. The only obvious difference is that the proportion of patients who has never worked living in urban area is much higher than living in rural area.

<<<<<<< HEAD

#### Relation between `stroke` and `avg_glucose_level`

According to common understandings, the risk for stroke is often associated with the average glucose level of an indivual. Thus, we chose to explore the relation between the outcome variable `strok` and `avg_glucose_level` using the boxplot below: 

=======
#### `age` and `bmi`
Among the relations between the numeric predictors, the correlation between age and bmi seems to be the strongest (most positive). Thus, we plotted the relation between `age` and `bmi` below: 
>>>>>>> a78639abc4fb18248797d564f7898ca053f63513
```{r}
# relation between `age` and `stroke`
ggplot(stroke_train, aes(avg_glucose_level, stroke)) + 
  geom_boxplot() + 
  labs(
     x = "Average Glucose Level",
     y = "Stroke",
     title = "The relationship between stroke and average glucose level"
  )
```
<br>

As shown, for observations with value `Yes` for `stroke` the interquantile range for the distribution of the average glucose level is much wider than the distribution of the average glucose level for level `No` of the `stroke`. Also, the median average glucose level for observations with `stroke` value `Yes` is larger than the the median average glucose level for observations with `stroke` value `No`. The distribution of average glucose level values is also more right-skewed for observations with `stroke` value `No`. For observations with `stroke` value `No`, average glucose level above around 180 are considered as outliers. We can infer from the plot that, for the sample population used for collecting this data, individuals with average glucose level above 125 are more likely to get stroke.



## Conclusion and Implications